Immutable Source-Mapped Abstract Syntax Tree:\ \\ A Design Pattern for Refactoring Engine APIs

### 2016-07-13 paper.tex
Fixed copyright footer
Prepare for TPLoP revisions
Updated Problems section to use reverse conditional.

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

\begin{abstract}
Many automated refactoring tools provide an extensibility mechanisms that
allows third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  In particular, these trees tend to have an
immutable structure, and they maintain very detailed information about the
source code, including exact character positions.  This pattern is used in the
Eclipse Java Development Tools' refactoring API as well as the Microsoft
``Roslyn'' CTP, among other uses.
\end{abstract}

@@

\copyr{PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
held by the author(s). ACM 978-1-4503-0107-7}

@@

Although you have an API for contributing refactorings/transformations, and you
have provided the ability to make string-like edits to source code, this is a
very primitive API that eschews any real knowledge about the syntactic
structure of the language being refactored.

Consider a simple Java refactoring that adds braces to the bodies of
\textit{if} statements, if they do not already exist.  For example, it would
make the following change.

\begin{center}
\vspace*{1em}
\begin{tabular}{ccc}
\begin{minipage}{2.25in}
\texttt{%
                if (a < b) \\
\hspace*{1.5em}   if (c < d) \\
\hspace*{3.0em}     System.out.println("a"); \\
\hspace*{1.5em}   else \\
\hspace*{3.0em}     System.out.println("b");} \\
 \\
\end{minipage}
& $\Longrightarrow$ &
\begin{minipage}{2.25in}
\texttt{%
                if (a < b) \textbf{\{} \\
\hspace*{1.5em}   if (c < d) \textbf{\{} \\
\hspace*{3.0em}     System.out.println("a"); \\
\hspace*{1.5em}   \textbf{\}} else \textbf{\{} \\
\hspace*{3.0em}     System.out.println("b"); \\
\hspace*{1.5em}   \textbf{\}} \\
                \textbf{\}}}
\end{minipage}
\end{tabular}
\vspace*{1em}
\end{center}

Even this simple refactoring requires nontrivial knowledge of Java's syntax,
including how the dangling \textit{else} is matched with a corresponding
\textit{if}.

How can the refactoring engine provide an API that exposes the syntactic
structure of programs in a way that is useful for implementing refactorings?

@@

\item The API must expose fine-grained details about the syntax of a program.
For example, the preceding example requires being able to determine the
presence or absence of braces \ttt{\{} \ttt{\}}, even though they are
semantically irrelevant.

>>>>

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY and CARL R.\ WORLEY \affil{Auburn University}}

\begin{abstract}
Many automated refactoring tools provide extensibility mechanisms that
allow third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  In particular, these trees tend to have an
immutable structure, and they maintain very detailed information about the
source code, including exact character positions.  This pattern is used in the
Eclipse Java Development Tools' refactoring API as well as the Microsoft
``Roslyn'' CTP, among other uses.
\end{abstract}

@@

\copyr{}%PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
%held by the author(s). ACM 978-1-4503-0107-7}

@@

Although you have an API for contributing refactorings/transformations, it is 
very primitive, only capable of string-like manipulations. This API lacks any
sophisticated knowledge of the text being edited.

Consider a simple Java refactoring that reverses the conditional meaning of
\textit{if} statements. For example, it would make the following change.

\begin{center}
\vspace*{1em}
\begin{tabular}{ccc}
\begin{minipage}{2.25in}
\texttt{%
                  if (!(a < b)) \textbf{\{}\\
\hspace*{1.5em}     System.out.println("a"); \\
                  \textbf{\}} else \textbf{\{}\\
\hspace*{1.5em}     System.out.println("b"); \\
                  \textbf{\}}} 
 \\
\end{minipage}
& $\Longrightarrow$ &
\begin{minipage}{2.25in}
\texttt{%
                  if (a < b) \textbf{\{} \\
\hspace*{1.5em}     System.out.println("b"); \\
                  \textbf{\}} else \textbf{\{} \\
\hspace*{1.5em}     System.out.println("a"); \\
                  \textbf{\}} \\
                }
\end{minipage}
\end{tabular}
\vspace*{1em}
\end{center}

To perform this refactoring, the conditional expression must be reversed in
meaning and the contents of the \textit{if} and \textit{else} branches must be
interchanged. Changing the condition requires removing the \textit{not} and the 
redundant parentheses. The former is simple enough, but the latter requires 
knowledge of matching parentheses. Interchanging the branches similarly requires 
knowledge of the matching braces, the location of the \textit{if} and dangling 
\textit{else}, and the presence or absence of any statements that alter flow of 
control such as \textit{goto}. In essence, knowledge of the program's syntax is 
required.

\textit{How can the refactoring engine provide an API that exposes the syntactic
structure of programs in a way that is useful for implementing refactorings?}

@@

\item The API must expose fine-grained details about the syntax of a program.
For example, the preceding example requires being able to determine the
exact location of parentheses \ttt{(} \ttt{)}.
